---
title: "Review for Exam 1"
author: "Andriana Manousidaki"
date: "2023-10-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Regression
The body weights of the chickens were measured at birth and every second day thereafter until day 20. They were also measured on day 21. There were four groups on chicks on different protein diets,we will focus only on 2 groups.
```{r}
path = 'https://vincentarelbundock.github.io/Rdatasets/csv/datasets/ChickWeight.csv'
chicken_weights <- read.csv(path, row.names = 1)
chicks=chicken_weights[chicken_weights$Diet%in%c(1,2),]
head(chicks)

```


**Fit a linear model for weight as a function of time and diet. Consider the possiblity that diet may also affect time.**

**Does diet affect weight?**

**Provide CI for the coefficient of time**



## Splines

**Continuing from the previous problem: A researcher wants to learn whether we have evidence of non-linear growth under Diet 1.Using the data for Diet 1 only, test for non-linearity using a cubic B-spline with 4 DF.**

```{r}
library(splines)
DATA=chicks[chicks$Diet==1,]
H0=lm(weight~Time,data=DATA)
HA=lm(weight~bs(Time,intercept=FALSE,df=4,degree=3),data=DATA)
anova(H0,HA)
```

## Maximum Likelihood Estimation 


** Find the mle of the parameters of beta distribution using optim().Information about the beta distribution can be find in https://en.wikipedia.org/wiki/Beta_distribution***
```{r}
y=rbeta(n=500, shape1=10, shape2=4)

```


##Other topics: Logistic regression, dimension reduction with svd.

