---
title: "Inclass 7"
author: "Andriana Manousidaki"
date: "2023-10-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1:Cholesky decomposition. 

Use the code below to generate a system of 5 equations on 5 unknowns (**Cb=r**). 
Solve the system using the Cholesky decomposition and `lm(y~X-1)`.
Use the function system.time() to compare the time needed to solve this system of equations with lm, Cholesky and Gauss-Seidel.

```{r}
 n=500
 p=5
 X=matrix(nrow=n,ncol=p,data=rnorm(n*p))
 y=X[,3]-X[,5]+rnorm(n)
 
 C=crossprod(X)
 r=crossprod(X,y)

```


## Q2:SVD. 

Below you can find how to read a data set of the gene expression data (columns) of Europeans and African ancestry people(row). Use SVD of X after scaling X, to reduce the deimension of the data set to two. Using the new dimension make a plot that illustrates the separation of the subjects to two groups.

```{r}
X=read.csv('https://www.dropbox.com/scl/fi/x1ixoaps699baao6xi3ye/GENOS_TWO_GROUPS.csv?rlkey=oirqands5fisg148o9xuadh97&dl=1',header=TRUE,row.names=1)

X=as.matrix(X)
dim(X)
X[1:10,1:3]


```
## Q3:Non linear regression for median value of houses in suburbs of Boston.

Using the Boston data of ISLR2 package, we want to asnwer the question, "Is there evidence of non-linear association between  the response(medv) any of the predictors(rest of variables)?" 

To answer this question, for each
predictor X, fit a model of the form Y = beta0 + beta1*X + beta2*X2 + beta3*X3 + e


```{r}
library(ISLR2)
head(Boston)

```