---
title: "Multiple Linear Regression"
author: "Andriana Manousidaki"
date: "2023-09-20"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data to R

```{r}
Gout = read.table('https://raw.githubusercontent.com/gdlc/STAT_COMP/master/DATA/goutData.txt', header = TRUE)

head(Gout)
```

## Explore the data

```{r}
table(Gout$sex)
table(Gout$race)

```

## Explore the model su~sex+race+age - Make the categorical variables numerical and create the incidence matrix

```{r}
X = model.matrix(~sex+race+age, data = Gout)

# Which levels are equal to 1

Gout[1:5,]
X[1:5,]
class(X[,1])
```


## The ordinary least square estimator $\hat{\beta}$

Get the estimates of the coefficients
```{r}
fm = lm(su ~ sex +race +age, data = Gout)
coef(fm)
```

Write the equation of the regression line 

```{r}

```

Make predictions predict(object, newdata)

```{r}
predict(fm)
```
Compare the predicted value for a Male, age  64, race black and a female, age  64, race black