---
title: "HW1_sol"
output: html_document
date: "2024-10-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Homework 1

Submit to (D2L -> Assessments -> Assignments). The deadline is Oct 4 (Friday) 10:00am. 

Using the Abalone dataset:

```{r}
abalone = read.table('https://raw.githubusercontent.com/gdlc/STAT_COMP/master/DATA/abalone.data', header = TRUE, sep=',')
```

**1)** Fit a linear model of the form  `Rings ~ .`, report your results, and summarize (in no more than three sentences) your conclusions.

```{r}
Q1.1 <- lm(Rings ~ ., data = abalone)
summary(Q1.1)
```

- Whether the abalone is infant or not significantly affect the number of rings. 
- Whether an adult abalone is male or female does not significantly affect the number of rings.
- Length is not a significant factor.
- All the other factors are significant.

**2)** Consider now expanding the model to include Length-by-Sex interactions. 

  - Fit the model with the interaction term, report your results and conclusions.
  
```{r}
Q1.2 <- lm(Rings ~ . + Sex*Length, data = abalone)
summary(Q1.2)
```  

- All factors are significant.
- The interaction between sex and length is also important.

**3)** Consider now testing the hypothesis that Sex has **any** effect on Rings (it could be an effect dependent on Length or independent of it) versus the null that states that Sex has no effect on Rings. 

  - Describe the null and the alternative hypothesis: H0: $\beta_{SexI}=\beta_{SexM}=\beta_{SexI:Length}=\beta_{SexM:Length}=0$. H1: at least one of the above four $\beta$s is nonzero.

  - Test the null using `anova()`, and
  - Summarize your findings.

```{r}
Q1.3.1 <- lm(Rings ~ . + Sex*Length, data = abalone)
Q1.3.0 <- lm(Rings ~ . -Sex , data = abalone)
print(anova(Q1.3.0,Q1.3.1))
```  
In the ANOVA, the p-value is almost zero, so we reject H0, and Sex has effect on Rings.

**4) Reproducing the results of the F-test**:  

  - Review the F-statistic in the [class notes](https://github.com/gdlc/STAT_COMP/blob/master/HANDOUTS/OLS.pdf) and

  - Develop a function that takes as input two `lm` objects and return a table identical to the one produced by `anova()`. 
  - Test your function using the H0 and Ha you used in Q3.


```{r}
 myANOVA=function(fm0,fm1){
   df0=length(coef(fm0))
   df1=length(coef(fm1))
   n=length(residuals(fm1))
   RSS0=sum(residuals(fm0)^2)   
   RSS1=sum(residuals(fm1)^2)  
   
   FSTAT=((RSS0-RSS1)/(df1-df0))/(RSS1/(n-df1))
   
   ANS=cbind('RES.DF'=c(n-df0,n-df1),'RSS'=c(RSS0,RSS1),DF=c(NA,df1-df0),
            'Sum of Sq'=c(NA,RSS0-RSS1),
            'F'=c(NA,FSTAT),
            'pValue'=c(NA,pf(lower.tail=FALSE,df1=df1-df0,df2=n-df1,q=FSTAT)))
   return(ANS)
 }
```


**Comparison**

```{r}
 anova(Q1.3.0,Q1.3.1)
 myANOVA(Q1.3.0,Q1.3.1)

```


**5)** Wald's test

Like the F-test, Wald's test can also be used for tests involving 1 or more than 1 df. The test can be used with any null that can be expressed in linear form. The general form of the test is as follows:

  - **Ha**: **y=Xb+e** (for this case use your Ha of Q3). Here, **y** is a nx1 vector (the *response*), **X** is an nxp incidence matrix for the pxy vector of effects **b**, and **e** is an nx1 error vector.
  - **H0**:  **Tb=a**, where  **T** is a contrast matrix of dimensions qxp, and **a** is a qx1 vector (often **a=0**).

The covariance matrix of the contrast ($\mathbf{\hat{d}=T\hat{b}}$) is $Cov(\mathbf{\hat{d}})=\mathbf{TCov(\hat{b})T'}=\mathbf{S}$, where $Cov(\hat{b})$ is the (co)variance matrix of estimates (Hint: use vcov(fm) to obtain it, here fm is the fitted alternative hypothesis). Note: here $\mathbf{\hat{b}}$ is the OLS estimate of **b** from Ha.

Because of the CLT, in large samples, $\mathbf{\hat{d}=T}\hat{\mathbf{b}}$  follows a multivariate normal distribution with (co)variance matrix **S**. Therefore, under the null, $(\mathbf{\hat{d}-a})'\mathbf{S}^{-1}(\mathbf{\hat{d}-a})$ follows a chi-square distribution with df equal to the rank of **T**.

  - Create a function that Implement Wald's test (your function should take a fitted model, representing Ha, and a matrix of contrasts (T). The function should return the test-statistic, test DF, and the p-value.
  
   - Test you function for the test in 3, compare your p-value with that of the F-test.

```{r}
 WTest=function(fm,T){
   DF=min(ncol(T),nrow(T))
   bHat=coef(fm)
   VCOV=vcov(fm)
   
   print(dim(T))
   print(bHat)
   
   if(ncol(T)!=length(bHat)){ stop('The number of columns of T must be equal to the number of parameters in the model')}
   dHat=T%*%bHat
   V=T%*%VCOV%*%t(T)
   SS=t(dHat)%*%solve(V)%*%dHat
   pvalue=pchisq(q=SS,df=DF,lower.tail=FALSE)
   return(c('df'=DF,'Chisq'=SS,'pvalue'=pvalue))
 }
 
 anova(Q1.3.0,Q1.3.1)
 T=rbind(c(0,1,0,0,0,0,0,0,0,0,0,0),
         c(0,0,1,0,0,0,0,0,0,0,0,0),
         c(0,0,0,0,0,0,0,0,0,0,1,0),
         c(0,0,0,0,0,0,0,0,0,0,0,1))

 WTest(Q1.3.1,T)

```
